---
title: "Clustering_Final"
author: "Evan Batzer"
date: "September 3, 2018"
output: html_document
---

```{r, include = FALSE}
library(tidyverse); library(testthat); library(vegan); library(msm); library(NbClust)
library(ggplot2); library(cluster); library(mclust); library(indicspecies)

# Cover data (wide format)
cover <- read.csv("../data/comdata_wideformat.csv", 
                  header = TRUE, 
                  stringsAsFactors = FALSE)
group_percents = cover[,8:19]
cover = cover[,-(8:19)]
comp.attr <- cover[,1:15]
comp.cover <- cover[,15:ncol(cover)]

# Treatment labels and group structure
trtlabels <- read.csv("../data/trtlabels.csv",
          header = TRUE,
          stringsAsFactors = FALSE)

# Species treatments
waps.specs <- read.csv("../data/WAPSspecies.csv",
                       header = TRUE,
                       stringsAsFactors = FALSE)

# Climate data
climate <- read.csv("../data/cimis_monthly.csv",
                       header = TRUE,
                       stringsAsFactors = FALSE)

```

# Clustering Communities - How best to define different states?

There are a number of different algorithms that can be used in cluster analysis, 
ranging from hierarchical, agglomerative methods that have been (in my
impression) a mainstay in Ecology, to more computationally intensive or statistically
nuanced methods, such as K-means (and its variations) or model-based clustering.

Based on my reading, there is no simple way to determine what type of algorithm
is best suited to your needs, or how many clusters are appropriate. According
to Legendre and Legendre (2008), all methods are valid, save for TWINSPAN, which
is cautioned against. If using K-means or hierarchical methods, there are some 
heuristics that can be used to identify the appropriate amount of clusters 
needed in the data. Model-based clustering is more complicated, but offers 
standard metrics to compare model fits -- typically, Bayesian Information 
Criterion (BIC). 

I think the best way to choose an appropriate classification algorithm is to 
explore a number of possibilities. Stein et al. (2016) and Bagchi et al. (2013) 
both employ hierarchical clustering algorithms - flexible beta and Ward's D, 
respectively - but we may have better luck using a method like K-means or 
one of the model-based clusters.

Some key considerations:

* __How to transform data prior to analysis?__ I think in most cases, focusing on the
proportional abundance of species in each sample makes the most sense, but there
may be some validity to using one of the other recommended transformations, such 
as Hellinger or Chi-Square.

* __What distance metric to use?__ Most ecological data utilizes Bray-Curtis or 
Euclidean distances between communities with abundance data. Bray-Curtis has a 
number of poor mathematical properties, however, that may make its use 
problematic in some circumstances. Horn and Morisita-Horn indices may be better,
but are more suited to individual counts an are less often used in circumstances
with cover data.

* __What species to include in analyses?__ When trying to cluster communities,
most analyses are conducted on either a relevant subset of species or a modified 
dataset. Stein et al. (2016) combine all of their weedy species / forbs into a 
single column called "forbs", while Bagchi et al. (2013) focus on species that 
occur in at least 2% of all samples taken. There are a lot of weedy species in 
the WAPS data that could be ignored in clustering as "noise".


```{r, echo = FALSE}
selected.trts <-c("annuals", "WAPS", "Natives", "natives+WAPS+annuals", "annuals+natives", "WAPS+annuals", "WAPS+natives")
selected.clips <- c("none")
selected.fert <- c("none")
selected.water <- c("control", "wet", "latewet", "dry") 

comp.subset = comp.cover[comp.attr$water %in% selected.water &
             comp.attr$sp.trt %in% selected.trts &
             comp.attr$clip %in% selected.clips &
             comp.attr$fertilization %in% selected.fert,]

print("Total Cover in Dataset:")
specsbycover = sort(round(colSums(comp.subset),2), decreasing = TRUE)
specsbycover

print("Plot Occupancy in Dataset:")
specsbyfreq = sort(round(colSums(decostand(comp.subset, method = "pa")) / nrow(comp.subset), 2), decreasing = TRUE)
specsbyfreq
```
* __When to start clustering?__ The seeded communities contain a number of species 
that are not well-represented in the data after the first year of planting. While
this is interesting information, it can be problematic for clustering, as it may make
the first year of data really distinct, almost a separate cluster in itself. My
guess is that Stein et al. (2016) found that this was the case with their data - 
they had 5 years of data in their study, but appeared to only run classification
and NMDS on the last 4 years. 

```{r, echo = FALSE}
# What treatments and years should be selected?
selected.trts <-c("annuals", "WAPS", "Natives", "natives+WAPS+annuals", "annuals+natives", "WAPS+annuals", "WAPS+natives")
selected.clips <- c("none")
selected.fert <- c("none")
selected.water <- c("control") 
selected.years <- c(2008:2018)

# Subsetting cover matrix
ord.mat <- comp.cover[comp.attr$water %in% selected.water &
                     comp.attr$sp.trt %in% selected.trts &
                     comp.attr$clip %in% selected.clips &
                     comp.attr$fertilization %in% selected.fert &
                      comp.attr$year %in% selected.years,]

# Subsetting attribute matrix
ord.att <- comp.attr[comp.attr$water %in% selected.water &
                     comp.attr$sp.trt %in% selected.trts &
                     comp.attr$clip %in% selected.clips &
                     comp.attr$fertilization %in% selected.fert &
                      comp.attr$year %in% selected.years,]

# Checking that these datasets are the same size
expect_true(nrow(ord.att) == nrow(ord.mat))

# Filtering species
totalcov = sum(ord.mat)
toselect = waps.specs$specname[waps.specs$group %in% c("annual", "natives", "waps")]
ord.mat = ord.mat %>% select(toselect)
cat("subsetcov =", sum(ord.mat) / totalcov)

# A number of rows have  no recorded live plants (mostly dry treatments?)
ord.att <- ord.att[rowSums(ord.mat) != 0, ]
ord.mat <- ord.mat[rowSums(ord.mat) != 0, ]
```

# Visualizing Data:

```{r}
# Constructing distance matrix and running NMDS
WAPS.mds <- metaMDS(decostand(ord.mat, "total"), distance = "bray", try = 1, trymax = 5, trace = F, autotransform = TRUE)
MDS_xy <- bind_cols(data.frame(ord.att), data.frame(X = WAPS.mds$points[,1], Y = WAPS.mds$points[,2]))

# Plotting figure
ggplot(MDS_xy, aes(X1, Y)) + 
  geom_point(aes(color = as.factor(year))) + 
  theme_bw() +
  geom_text(data = data.frame(X1 = Inf, Y = -Inf), 
            hjust = 1, vjust = -1, 
            label = paste("Stress =", round(WAPS.mds$stress, 2)))
```

# K-means Clustering

## Selecting an appropriate number of clusters

```{r, include = FALSE}
counter = 1
groupdiss <- c()
groupsil <- c()

distmat <- vegdist(decostand(ord.mat, "total"), distance = "bray")

for(i in 2:10){
  tempclust = pam(distmat, i)
  groupdiss[counter] = mean(tempclust$clusinfo[,3])
  groupsil[counter] = tempclust$silinfo$avg.width
  counter <- counter + 1
}

data.frame(k = c(2:10), dis = groupdiss, sil = groupsil) %>% 
  ggplot(aes(x = k,
             y = groupsil,
             fill = groupsil == max(groupsil))) +
  geom_point(size = 3, pch = 21)

data.frame(k = c(2:10), dis = groupdiss, sil = groupsil) %>% 
  ggplot(aes(x = k,
             y = groupdiss,
             fill = groupsil == max(groupsil))) +
  geom_point(size = 3, pch = 21)
```

When comparing a number of different criteria, 4 seems to be the best number of clusters to use
(see best.nc index) -- the nbclust() package in R suggests using a battery of different 
metrics and seeing if some cluster number appears as a consistent good performer.
In this case, 4 clusters is selected as the best number in 4 of the 8 metrics,
and consistently performs well (usually in the top 2-4 of the 10) for the rest.

Apparently, 7 clusters also seems like a good second choice by these metrics,
though that number seems too high to be relevant given the number of replicates 
we have, as well as my intuition for the system as a whole. 

```{r, warning = FALSE}
NbClust(decostand(ord.mat, method = "total"), min.nc = 2, max.nc = 10, method = "kmeans", 
        index = c("hartigan", "ch", "beale", "kl", "cindex", "db", "silhouette", "duda"))[1:2]
```

## Visualizing how the clustering algorithm splits up the data

```{r}
distmat <- vegdist(decostand(ord.mat, "total"), "bray")
k.cluster = pam(distmat, 4)
k.cluster$objective
MDS_xy$clust = k.cluster$clustering

ggplot(MDS_xy, aes(X1, Y, color = as.factor(clust))) + 
  geom_point() + 
  theme_bw() +
  geom_text(data = data.frame(X1 = Inf, Y = -Inf), 
            aes(color = NULL), hjust = 1, vjust = -1, 
            label = paste("Stress =", round(WAPS.mds$stress, 2)))

cluscounts = bind_cols(ord.att, data.frame(cluster = k.cluster$clustering)) %>% group_by(year, cluster) %>% summarise(count = n())
  
ggplot(cluscounts, aes(x = factor(year), y = count, fill = factor(cluster))) + geom_bar(stat = "identity")
```

# Describing clusters:

## Indicator species analysis:

```{r}
IS = multipatt(decostand(ord.mat, method = "total"), k.cluster$clustering, duleg = TRUE, control = how(nperm=999))
summary(IS, alpha = .05)
```

Visualizing differences

```{r}
piecharts = bind_cols(data.frame(cluster = k.cluster$clustering), ord.mat)
piecharts$cluster[piecharts$cluster == 1] = "Natives"
piecharts$cluster[piecharts$cluster == 2] = "Wet Annuals"
piecharts$cluster[piecharts$cluster == 3] = "WAPS"
piecharts$cluster[piecharts$cluster == 4] = "Dry Annuals"

piecharts %>%
  gather(key = "species", value = "cover", -cluster) %>%
  group_by(cluster, species) %>%
  summarise(cover = mean(cover)) %>%
  ggplot(aes(x = "", y = cover, fill = species))+
  geom_bar(stat = "identity", position = position_fill(), color = "black") +
  # geom_text(aes(label = Cnt), position = position_fill(vjust = 0.5)) +
  coord_polar(theta = "y") +
  facet_wrap(~cluster) +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank(),  axis.ticks = element_blank(),
        panel.grid  = element_blank(), axis.text = element_blank(), legend.title = element_blank()) +
  theme(legend.position='bottom') + guides(fill=guide_legend(nrow=2,byrow=TRUE)) +
  theme(strip.text.x = element_text(size = 20, face = "bold"))

ggsave("../figures/clustercomps.pdf", height = 10, width = 10)
```

# Adding indicator species vectors to NMDS plot

```{r}

specvecs = IS$sign %>% 
  rownames_to_column() %>% 
  left_join(data.frame(WAPS.mds$species) %>% rownames_to_column()) %>% 
  filter(p.value < .05)

specvecs$rowname <- str_replace(specvecs$rowname, 
                                pattern = c("avena", "bro.hordeaceous", "bro.diandrus", "lolium", "trifolium", "bro.carinatus", "elymus", "leymus", "lotus", "nassella", "aegilops", "taeniatherum"), 
                                replace = c("A. fatua", "B. hordeaceous", "B. diandrus", "F. perennis", "T. subterreaneum", "B. carinatus", "E. glaucus", "E. triticoides", "A. americanus", "S. pulchra", "A. triuncialis", "E. caput-medusae"))
```

```{r}
ggplot(MDS_xy, aes(X1, Y)) + 
  geom_point(aes(color = as.factor(clust),
                 shape = as.factor(clust))) + 
  theme_bw() +
  geom_text(data = data.frame(X1 = Inf, Y = -Inf), 
            aes(color = NULL), hjust = 1, vjust = -1, 
            label = paste("Stress =", round(WAPS.mds$stress, 2)), show.legend = FALSE) +
  geom_segment(data = specvecs,
               aes(x = 0,
                   y = 0, 
                   xend = MDS1,
                   yend = MDS2,
                   color = NULL),
               arrow = arrow(length = unit(0.01, "npc"), type = "closed"), show.legend = FALSE) + 
  geom_text(data = specvecs,
            aes(x = MDS1, 
                y = MDS2,
                color = NULL,
                label = rowname),
            nudge_x = 0.05,
            nudge_y = 0.03,
            show.legend = FALSE) +
  guides(color=guide_legend(title="Cluster ID"),
         shape = guide_legend(title="Cluster ID")) +
  xlab("NMDS1") +
  ylab("NMDS2")
```



# Plotting transitions between clusters over time

```{r}
clustdat <- bind_cols(ord.att, data.frame(cluster = k.cluster$clustering)) %>%
  dplyr::arrange(sp.trt, plot)

# Assigning states
clustdat$cluster[clustdat$cluster == 1] = "Cluster 1"
clustdat$cluster[clustdat$cluster == 2] = "Cluster 2"
clustdat$cluster[clustdat$cluster == 3] = "Cluster 3"
clustdat$cluster[clustdat$cluster == 4] = "Cluster 4"

clustdat %>% group_by(plot) %>% summarise(count = n()) %>% arrange(count)

clustdat 
```

```{r}
clustdat[clustdat$plot == 205,]
```

```{r}
# Renaming Treatments

searchvals = c(unique(clustdat$sp.trt))

toreplace = c("1) Exotic", "2) Exotic + Native", "7) Native", "3) Exotic + Invasive + Native", "5) Invasives", "4) Exotic + Invasive", "6) Invasive + Native")

for(i in 1:length(searchvals)){
  clustdat$sp.trt[clustdat$sp.trt == searchvals[i]] = toreplace[i]
}

# Plotting change in assignment over time
clustdat$plot <- factor(clustdat$plot, levels = unique(clustdat$plot))

clustdat = clustdat %>% arrange(sp.trt, plot)

fixplots <- function(x){
  newvec <- c()
  for(i in 1:length(x)){
    if(i == 1){
      newvec[i] = 1
    }
    else{
      if(x[i] == x[i - 1]){
        newvec[i] = newvec[i - 1]
      }else{
        newvec[i] = newvec[i - 1] + 1
      }
    }
  }
  
  return(newvec)
}

clustdat$plot <- fixplots(clustdat$plot)
```

```{r}

gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

cols = gg_color_hue(4)

clustdat %>% ggplot(aes(x = year,
           y = as.numeric(plot) - .5)) +
  theme_bw() + 
  geom_tile(color = "black", aes(fill = cluster)) +
  geom_hline(yintercept = c(seq(0, 56, by = 8)), size = 2) +
  geom_text(label = unique(clustdat$sp.trt),
            data = data.frame(year = rep(2004, 7),
                              plot = seq(4, 52, by = 8)),
            fontface = "bold") +
  scale_fill_manual(name = 'State Assignment', 
         values =c("Cluster 1" = cols[1], 
                   "Cluster 2" = cols[2], 
                   "Cluster 3" = cols[3], 
                   "Cluster 4" = cols[4]),
         labels = c('E. glaucus - S. pulchra',
                    'F. perennis - B. hordeaceous', 
                    'E. caput-medusae - A. triuncialis',
                    'A. fatua - B. diandrus')) +
  xlim(2001, 2019) +
  geom_vline(xintercept = 2008.5, color = "black", size = 2) +
  geom_vline(xintercept = 2007.5, color = "black", size = 2) +
  ylab(NULL) +
  xlab("Year") + 
  ylim(0, 57) +
  theme(axis.text.y = element_blank())

ggsave("../figures/TrackingStates.pdf", height = 6, width = 10)

clustdat %>% ggplot(aes(x = year,
           y = as.numeric(plot) - .5)) +
  theme_bw() + 
  geom_point(aes(shape = cluster, fill = cluster), size =2) +
  # geom_hline(yintercept = c(seq(0, 56, by = 8)), size = 2) +
  geom_text(label = unique(clustdat$sp.trt),
            data = data.frame(year = rep(1999, 7),
                              plot = seq(4, 52, by = 8)),
            fontface = "bold") +
  scale_fill_manual(name = 'State Assignment', 
         values =c("Cluster 1" = cols[1], 
                   "Cluster 2" = cols[2], 
                   "Cluster 3" = cols[3], 
                   "Cluster 4" = cols[4]),
         labels = c('E. glaucus - S. pulchra',
                    'F. perennis - B. hordeaceous', 
                    'E. caput-medusae - A. triuncialis',
                    'A. fatua - B. diandrus')) +
    scale_shape_manual(name = 'State Assignment', 
         values =c("Cluster 1" = 21, 
                   "Cluster 2" = 22, 
                   "Cluster 3" = 23, 
                   "Cluster 4" = 24),
         labels = c('E. glaucus - S. pulchra',
                    'F. perennis - B. hordeaceous', 
                    'E. caput-medusae - A. triuncialis',
                    'A. fatua - B. diandrus')) +
  xlim(1990, 2019) +
  geom_vline(xintercept = 2008.5, color = "black", size = 1) +
  geom_vline(xintercept = 2007.5, color = "black", size = 1) +
  ylab(NULL) +
  xlab("Year") + 
  ylim(0, 57) +
  theme(axis.text.y = element_blank())

ggsave("../figures/TrackingStates_alternate.pdf", height = 6, width = 10)

```

# Frequency of transitions

How often do transitions occur?

```{r}
next_state = clustdat %>% mutate(year = year + 1) %>% rename(paststate = cluster)
transitions = left_join(clustdat, next_state, by = c("plot", "year")) %>% 
  mutate(transition = as.numeric(!cluster == paststate))

transto = transitions %>% filter(transition == 1) %>% group_by(cluster, year) %>% 
  summarise(total = sum(transition))

climate$month = gsub("-.*", "", climate$Month.Year)

climsum = climate %>% group_by(Water.Year) %>% 
  summarise(total = sum(Total.Precip..mm.), 
            early = sum(Total.Precip..mm.[month %in% c("Oct", "Nov", "Dec")]),
            mid = sum(Total.Precip..mm.[month %in% c("Jan", "Feb", "March")]),
            late = sum(Total.Precip..mm.[month %in% c("April", "May")]))

transitions %>% group_by(year) %>% filter(year != 2008) %>%
  summarise(trans = sum(na.omit(transition))) %>%
  ggplot() +
  geom_bar(aes(x = year,
             y = trans / 56),
           stat = "identity",
           color = "black", 
           fill = "forestgreen") +
  geom_line(data = climsum %>% filter(Water.Year > 2007),
            aes(x = Water.Year ,
                y = total / max(na.omit(total))))

transto %>%
  ggplot() +
  geom_bar(aes(x = year,
             y = total / 56,
             fill = cluster),
           stat = "identity",
           color = "black")+
  geom_line(data = climsum %>% filter(Water.Year > 2007),
            aes(x = Water.Year,
                y = total / max(na.omit(total)) + 1)) +
    geom_line(data = climsum %>% filter(Water.Year > 2007),
            aes(x = Water.Year,
                y = early / max(na.omit(total)) + 1),
            color = "blue") +
    geom_line(data = climsum %>% filter(Water.Year > 2007),
            aes(x = Water.Year,
                y = mid / max(na.omit(total)) + 1),
            color = "orange") +
    geom_line(data = climsum %>% filter(Water.Year > 2007),
            aes(x = Water.Year,
                y = late / max(na.omit(total)) + 1),
            color = "red") +
  geom_hline(yintercept = 1)
```

# Writing new CSV with cluster data:

```{r}
clustdat <- bind_cols(ord.att, data.frame(cluster = k.cluster$clustering)) %>%
  dplyr::arrange(sp.trt, plot)

#write.csv("../data/WAPSclusters.csv", x = clustdat)
```

