---
title: "WAPS Data Cleaning and Validation"
output:
  html_document:
    df_print: paged
  prettydoc::html_pretty:
  highlight: github
---

This notebook serves as an introductory step in the analysis of WAPS data from 2008-2018, examining the interaction between planted taxa (priority effects) and environmental treatments (watering, fertilization, and clipping) on the maintenance of species diversity and compositional dissimilarity in California grasslands. 

## Initial data processing

Loaded files:

* WAPSthrough2017 - Dataframe of composition data (areal estimates of cover by species)
* trtlabels - Dataframe of treatment labels for each plot (planted taxa, treatments, blocking structure)
* WAPSspecies - Dataframe of project species and associated labels/groups

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Cleaning up WAPS data
library(dplyr);library(tidyr);library(vegan);library(testthat)

# Set working directiory
setwd("C:/Users/ebatz/Box Sync/waps/evan-sarah analyses")

# Reading in datasets

# Species composition
waps.comp <- read.csv("./data/WAPSthrough2017.csv",
          header = TRUE,
          stringsAsFactors = FALSE)

# Treatment labels and group structure
trtlabels <- read.csv("./data/trtlabels.csv",
          header = TRUE,
          stringsAsFactors = FALSE)

# Species labels
waps.specs <- read.csv("./data/WAPSspecies.csv", 
                       stringsAsFactors = F)
```

### Data cleaning

There are a bunch of small inconsistencies that come with manual data entry that need to be corrected prior to analysis.

__Checklist:__  

* Convert dataset to wide (matrix format) __done__

* Standardize species names __done__

* Clean up year column __done__

* Assign NA cover values to 0 __done__

* Standardize treatment labels __done__

* Add "class" statistics to attribute table __done__

```{r,  echo = FALSE, message = FALSE, warning = FALSE}
# Gathers dataset to long format before cleaning up, creates species column
long.comp <- gather(waps.comp, key = "species", value = "cover", -c(1:7))

# Removes year labels from species column
spec <- gsub("X\\d\\d\\.*", "", long.comp$species)

# Grabs the first two digits from the species column to make a "year" column
yr <- gsub('\\.', "", long.comp$species)
yr <- gsub("X", "20", yr)
yr <- gsub("[[:alpha:]]+0?", "", yr)

# Reassigns species variable
long.comp$species <- spec

# Converts to lowercase
long.comp$species <- tolower(long.comp$species)

# Attaches year variable
long.comp <- cbind(long.comp, year = as.numeric(yr))

# Lowercase names for dataset
names(long.comp) <- c("plot", "block", "subblock",
                      "water", "spcomp", "fertilization",
                      "clipping", "species", "cover", "year")

# Assigns all NA cover values to zero
long.comp$cover[is.na(long.comp$cover)] <- 0
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Reassigns species names to standardized names in the waps.specs dataset
for(i in waps.specs$label){
  long.comp$species <- gsub(paste(i, "$", sep = ""), waps.specs$specname[waps.specs$label == i], long.comp$species)
}

# Removes spaces from species names
long.comp$spcomp <- gsub("[[:space:]]\\+[[:space:]]*", " \\+", long.comp$spcomp)

# Removes whitespace from treatment names
long.comp$spcomp <- gsub("[[:space:]]", "", long.comp$spcomp)
long.comp$water <- gsub("[[:space:]]", "", long.comp$water)

# Assigns all NAs to zero, merges duplicated species names
long.comp$cover <- as.numeric(long.comp$cover)
long.comp$cover[is.na(long.comp$cover)] <- 0
long.comp <- long.comp %>% group_by(plot, block, subblock, water, fertilization, species, year) %>% summarise(cover = sum(cover))

# Writing csv of long format cover data
write.csv(x = long.comp, file = "../data/comdata_longformat.csv", sep =",")
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Spreads dataset to wide format
comp.wide <- spread(long.comp, key = "species", value = "cover")
comp.attr <- select(comp.wide, one_of(c("year", waps.specs$specname[waps.specs$type == "class"])))
comp.wide <- ungroup(comp.wide) %>%  select(one_of(waps.specs$specname[waps.specs$type != "class"]))
comp.wide <- apply(comp.wide, 2, as.numeric)
comp.wide[is.na(comp.wide)] <- 0

# Merges with treatment labels
comp.attr <- left_join(comp.attr, trtlabels)
comp.attr$sp.trt <- gsub("[[:space:]]", "", comp.attr$sp.trt)

# Binding attribute columns with abundance matrix and writing new CSV
wide.comp <- cbind(data.frame(comp.attr), data.frame(comp.wide))

# Writing full wide csv
write.csv(x = wide.comp, file = "../data/comdata_wideformat.csv")
```

# Error checking:

### What species are present within the dataset:

There seem to be no obvious distinctions among the different groups in this dataset, but need to double-check with Val and David about manual data collection.  

Some things to consider:

__Distinctions between labels in the dataset:__ 

* elymus vs. leymus 
* other vs. other.weeds 
* thistle vs. unk.thistle vs. milk.thistle vs. italian.thistle
* tarweed vs. any other species?
* bare vs. disturbance

__Specific species IDs__

* What species is mustard? 
* What species is oats? 
* What species is tarweed? 
* What species is trifolium? 
* What species is big.leaf.dandelion? 

```{r}
unique(long.comp$species)
```

### How are cover values distributed? Are there outliers?

Just looking at the disribution of cover values, we see that a number of samples have no live cover (which may be expected), but the distribution is quite right-skewed.

```{r}
hist(rowSums(comp.wide))
```

__Which plots are responsible for this skewed data?__

This table seems to suggest that, in particular, some of the highest row values occur where species like Taeniatherum have really high cover values (>100). I suspect that these values should be changed from 235 to 23.5. 

Some other values are also really high - why might this be the case? Worth checking with Valerie and David.

```{r}
wide.comp[order(rowSums(comp.wide), decreasing = TRUE)[1:10],c(1:6, 26:ncol(wide.comp))]
```

__Unique cover values__
```{r}
sort(unique(c(comp.wide)))
```

__Total number of cover values > 100__ 
```{r}
sum(c(comp.wide) > 100)
```

__Changing these entries to 23.50 seems to solve the problem:__
```{r}
comp.wide[comp.wide == 235] = 23.50
hist(rowSums(comp.wide))
wide.comp <- cbind(data.frame(comp.attr), data.frame(comp.wide))
write.csv(x = wide.comp, file = "../data/comdata_wideformat.csv")
```

