---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
```

# Exploring precipitation patterns

What is the best way to explore community responses to variation in precipitation patterns?

Papers on the subject seem to suggest a number of different ways to subset climate
data to make sense of changes in community composition and productivity from year
to year. 

Some relevant citations and their approaches to using climate patterns:

* Dudney et al. (2016)

Included both current-year and previous-year rainfall effects divided into Fall
(October - November), Winter (December - February), and Spring (March - May). Significant
variables were current winter, current spring, previous fall, and previous spring. Coefficients
of similar magnitudes suggest that no one season is most responsible for change. 

* Adler and Levine (2007)

Included both current-year and previous-year rainfall effects, cumulatively by year.
Also included transformations of variables to better account for nonlinearity of effects;
current and last year richness / productivity was best explained by the inverse 
transformation of current and last year precipitation.

* Pitt and Heady (1978)

Included 76(!) variables in analysis -- mean, minimum, and maximum temperatures,
number of days receiving <6.35 mm rain, total precip, days below freezing,
drough (longest successive number of days receiving <1.27 mm of rain), and
germination date (number of days after 31 first storm of 19mm).

* Eviner (2015) has a general overview:

Plant production highest in years with high and steady rainfall between November
and February, especially when temperatures are high in that period. Site-specific.

High precipitation with warm temperatures in the Fall tend to favor annual grasses,
largely due to rapid germination in warm, moist conditions.

Fall rains followed by a prolonged fall or early-winter drought tend to favor 
forbs and legumes.

Prolonged mid-winter drouhgt tends to favor forbs, clovers, and perennial grasses.

Some potential variables of interest:

* 19mm (3/4") of an inch needed for grasses to germinate

Framework for how to measure climatic variables:

1. Define the start of each year's germinating season (rainfall in 24hrs > 19mm)

2. Calculate a series of potential variables:
          * Date of season start and precip
          * Mean precip for each 30-day period following the season start
          * Total number of days without precipitation in 30-day intervals
          * Maximum number of sequential days without precipitation

```{r, echo = FALSE, warning = FALSE, message=FALSE}
library(lubridate); library(tidyverse); library(tidyquant); library(msm)

clustdat <- read.csv("../data/WAPSclusters.csv", 
                     header = TRUE, 
                     stringsAsFactors = FALSE)

climate <- read.csv("../data/cimis_monthly_1980.csv",
                       header = TRUE,
                       stringsAsFactors = FALSE)

climate_daily <- read.csv("../data/cimis_daily.csv",
                       header = TRUE,
                       stringsAsFactors = FALSE)

climate_daily$Date = as.POSIXct(climate_daily$Date, format = "%m/%d/%Y")

```

# First date of sufficient germinating rains

Following figures show the first date when the total amount of precipitation exceeded 19mm in a rolling 7-day period. 

Was it really so late in 2012-2014?

```{r}
climate_daily %>% mutate(month = month(Date),
                         year = year(Date)) %>%
  arrange(Date) %>%
  mutate(precip_run = c(rep(0, 6), rollapply(Precip..mm., 7, sum))) %>%
  group_by(Water.Year) %>% 
  filter(precip_run > 19) %>%
  summarise(GermRain = min(Date))
```
 
# Average rain in each month

Following figures show average precipitation (in mm) for each month across the growing season. 

```{r}
raintotal = climate_daily %>% mutate(month = month(Date),
                         year = year(Date)) %>%
  group_by(Water.Year, month) %>%
  summarise(totalprecip = sum(na.omit(Precip..mm.)),
            year = unique(year)) %>%
  filter(month %in% c(11,12,1,2,3,4,5)) %>%
  mutate(date = as.POSIXct(paste(year, month,1,sep = "/"),
                           format = "%Y/%m/%d"))

raintotal %>% 
  filter(Water.Year > 2007) %>%
  ggplot(aes(x = date,
             y = totalprecip,
             fill = as.factor((month + 2) %% 12))) +
  geom_bar(stat = "identity") +
  viridis::scale_fill_viridis(discrete = TRUE,
                              labels = c("Nov", "Dec", "Jan",
                                         "Feb", "Mar", "Apr", "May")) +
  labs(fill = "Month", y = "Total Precip (mm)", x = "Year") +
  ggtitle("Precipitation over Time")
```

# CV of rainfall in each year

Another way to think about climatic stress would be to quantify how variable precipitation is in a given year -- does consistent precipitation generate a different community than more variable precip? Here, the measure of a coefficient of variation ($CV = \dfrac{stddev}{mean}$) might be the best measure of variability.

```{r}
precipCV = climate_daily %>% mutate(month = month(Date),
                         year = year(Date)) %>%
  filter(month %in% c(11,12,1)) %>%
  group_by(Water.Year) %>%
  summarise(pptCV = sd(na.omit(Precip..mm.))/
                                   mean(na.omit(Precip..mm.)))
precipCV %>% 
  filter(Water.Year > 2007) %>%
  ggplot(aes(x = Water.Year,
             y = pptCV,
             fill = pptCV)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c()
```

# Total number of days in each month with no rain (or less than 1mm?)

Following figures show the total number of days in each year where no rain is observed between November and March

```{r}
droughttotal = climate_daily %>% mutate(month = month(Date),
                         year = year(Date)) %>%
  group_by(Water.Year, month) %>%
  summarise(droughtdays = sum(na.omit(Precip..mm.) < 1)) %>%
  filter(month %in% c(11,12,1,2,3))

droughttotal %>% 
  group_by(Water.Year) %>% 
  summarise(Total = sum(droughtdays))
```

# Length of droughts in each water year

We can also try to quantify the number and duration of droughts within each year - I've written a function that produces a table of all droughts (consecutive days with less than a mm of rain), and highlighted the maximum drought lenght in each year.

```{r}
definedrought <- function(df){
  
  counter = 1
  output <- list()  
  dstart = NA
  dend = NA

  for(i in 1:nrow(df)){
    if(df$Precip..mm.[i] < 1 & is.na(dstart)){
      dstart = df$Date[i]
    }
    
    if(df$Precip..mm.[i] > 1 & !is.na(dstart)){
      dend = df$Date[i]
      output[[counter]] = data.frame(dstart, dend, duration = dend - dstart)
      dstart = NA
      dend = NA
      counter <- counter + 1
    }
  }
  
  return(bind_rows(output))
  
}

climate_daily = climate_daily[!is.na(climate_daily$Precip..mm.),]

droughtsummary = climate_daily %>% 
  mutate(month = month(Date),
         year = year(Date)) %>% 
  arrange(Date) %>%
  group_by(Water.Year) %>%
  filter(month %in% c(11,12,1,2,3)) %>%
  do(definedrought(.)) %>%
  mutate(duration = round(duration))

droughtsummary %>% 
  ggplot(aes(x = duration)) +
  geom_histogram()

cat("Longest Observed Winter Droughts:")
droughtsummary %>%
  group_by(Water.Year) %>%
  filter(duration == max(duration))
```

# SPEI - a standard drought severity metric used in many studies (Prugh et al. 2018, etc.)

From Prugh et al:

"We used the FAO-56 Penman–Monteith equation implemented in the ‘SPEI’ R package 
to calculate atmospheric evaporative demand (ETo) each month based on average 
daily minimum and maximum temperatures, average daily relative humidity, dew-point 
temperature, solar radiation and wind speed. SPEI was then calculated as precipitation 
minus ETo. SPEI can be calculated over different time intervals to measure short
- or long-term droughts. To understand aridity in April at the peak of primary 
production, we calculated 7-, 19- and 31-month SPEI values, which estimated water 
balance over 1, 2 and 3 growing seasons, respectively, before the current time point. 
We used the reverse of these SPEI values as a drought index, such that 
DIx=−1×SPEIx to calculate DI1, DI2 and DI3, respectively from 2000–2015."

So, this drought calculation can be performed across "sliding timescales", allowing 
us to make drought comparisons between 1, 2, and 3 + year intervals.

This paper makes use of the Penman-Montieth equation, which can be used to estimate
PET. This is a metric that is supported by the FAO as a standardized metric for comparison.

The SPEI package in R makes the implementation of the Penman-Monteith and SPEI 
index quite simple.

Relevant data provided in the CIMIS dataset include:

__MONTHLY INTERVALS!__

* Min temp
* Max temp
* Wind speed
* Dew point
* Solar radiation
* Vapor pressure deficit
* Average relative humidity

CIMIS actually uses this Penman-Monteith equation to estimate ETo, so we should be able to 

```{r}
climate$Date = as.POSIXct(climate$Month.Year, format = "%m/%d/%Y")
climate = climate[year(climate$Date) > 1982,]
```

```{r}
library(SPEI)
lag1_SPEI = spei(data = na.omit(climate$Total.Precip..mm. - climate$Total.ETo..mm.), scale = 8)
lag2_SPEI = spei(data = na.omit(climate$Total.Precip..mm. - climate$Total.ETo..mm.), scale = 20)
lag3_SPEI = spei(data = na.omit(climate$Total.Precip..mm. - climate$Total.ETo..mm.), scale = 32)
speifits = data.frame(year = sort(rep(1983:2018, 12)), 
                      month = rep(c(1:12), 12),
                      l1fit = c(lag1_SPEI$fitted, NA, NA),
                      l2fit = c(lag2_SPEI$fitted, NA, NA),
                      l3fit = c(lag3_SPEI$fitted, NA, NA))

write.csv("../data/SPEIfits.csv", x = speifits %>% filter(month == 5))

p1 = speifits %>% filter(month == 5) %>%
  gather(key = fit, value = spei, - year, - month) %>%
  ggplot(aes(x = year, y = spei)) +
  geom_line(aes(linetype = fit), size = 1.5) +
  ggtitle("SPEI") +
  ylab("SPEI Index") +
  xlab("Year") +
  scale_linetype_manual(values=c("solid", "twodash", "dotted"),
                        name = "Temporal Lag",
                        labels = c("1 Year", "2 Years", "3 Years")) + 
  theme_bw() +
  geom_hline(yintercept = 0, linetype = "dashed") + 
  scale_x_continuous(breaks = seq(2008, 2018, by = 2), limits = c(2008, 2018))

climate$wateryr = year(climate$Date)
climate$wateryr = if_else(month(climate$Date) > 9, climate$wateryr + 1, climate$wateryr)

climsum = climate %>% group_by(wateryr) %>%
  summarise(Total.Precip..mm. = sum(Total.Precip..mm.))

p2 = climsum %>%
  ggplot(aes(x = wateryr,
             y = Total.Precip..mm.)) +
  geom_line() +
  theme_bw() +
  ylab("Total Precip (mm)") +
  xlab("Water Year") +
  ggtitle("Precipitation")+
  geom_hline(yintercept = mean(na.omit(climsum$Total.Precip..mm.)), 
             linetype = "dashed") + 
  scale_x_continuous(breaks = seq(2008, 2018, by = 2), limits = c(2008, 2018))


library(gridExtra)
grobglob = gridExtra::grid.arrange(p2, p1, nrow = 1, widths = c(1.5,2))
ggsave(plot = grobglob, "../figures/precip_spei.pdf", width = 10, height = 5)
```



