---
title: "Introductory analysis of cover and diversity trends"
output: html_notebook
---

This notebook serves as an introductory step in the analysis of WAPS data from 2008-2018, examining the interaction between planted taxa (priority effects) and environmental treatments (watering, fertilization, and clipping) on the maintenance of species diversity and compositional dissimilarity in California grasslands. 

## Initial data processing

Loaded files:

* WAPSthrough2017 - Dataframe of composition data (areal estimates of cover by species)
* trtlabels - Dataframe of treatment labels for each plot (planted taxa, treatments, blocking structure)
* WAPSspecies - Dataframe of project species and associated labels/groups

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Cleaning up WAPS data
library(dplyr);library(tidyr);library(vegan);library(testthat)

# Set working directiory
setwd("C:/Users/ebatz/Box Sync/waps/evan-sarah analyses")

# Read in datasets

# Species composition
waps.comp <- read.csv("../data/WAPSthrough2017.csv",
          header = TRUE,
          stringsAsFactors = FALSE)

# Treatment labels and group structure
trtlabels <- read.csv("../data/trtlabels.csv",
          header = TRUE,
          stringsAsFactors = FALSE)

# Species labels
waps.specs <- read.csv("../data/WAPSspecies.csv", 
                       stringsAsFactors = F)
```

### Data cleaning

There are a bunch of small inconsistencies that come with manual data entry that need to be corrected prior to analysis.

__Checklist:__  

* Convert dataset to wide (matrix format) __done__

* Standardize species names __done__

* Clean up year column __done__

* Assign NA cover values to 0 __done__

* Standardize treatment labels __done__

* Add "class" statistics to attribute table __done__

```{r,  echo = FALSE, message = FALSE, warning = FALSE}
# Gathers dataset to long format before cleaning up, creates species column
long.comp <- gather(waps.comp, key = "species", value = "cover", -c(1:7))

# Removes year labels from species column
spec <- gsub("X\\d\\d\\.*", "", long.comp$species)

# Grabs the first two digits from the species column to make a "year" column
yr <- gsub('\\.', "", long.comp$species)
yr <- gsub("X", "20", yr)
yr <- gsub("[[:alpha:]]+0?", "", yr)

# Reassigns species variable
long.comp$species <- spec

# Attaches year variable
long.comp <- cbind(long.comp, year = as.numeric(yr))

# Lowercase names for dataset
names(long.comp) <- c("plot", "block", "subblock",
                      "water", "spcomp", "fertilization",
                      "clipping", "species", "cover", "year")

# Assigns all NA cover values to zero
long.comp$cover[is.na(long.comp$cover)] <- 0
```

```{r,  echo = FALSE, message = FALSE, warning = FALSE}
# Reassigns species names to standardized names in the waps.specs dataset
for(i in waps.specs$Species){
  long.comp$species <- gsub(paste(i, "$", sep = ""), waps.specs$specname[waps.specs$label == i], long.comp$species)
}

# Removes spaces from species names
long.comp$spcomp <- gsub("[[:space:]]\\+[[:space:]]*", " \\+", long.comp$spcomp)

# Removes whitespace from treatment names
long.comp$spcomp <- gsub("[[:space:]]", "", long.comp$spcomp)
long.comp$water <- gsub("[[:space:]]", "", long.comp$water)

# Assigns all NAs to zero, merges duplicated species names
long.comp$cover <- as.numeric(long.comp$cover)
long.comp$cover[is.na(long.comp$cover)] <- 0
long.comp <- long.comp %>% group_by(plot, block, subblock, water, fertilization, species, year) %>% summarise(cover = sum(cover))

# Spreads dataset to wide format
comp.wide <- spread(long.comp, key = "species", value = "cover")
comp.attr <- select(comp.wide, one_of(c("year", waps.specs$Newval[waps.specs$Type == "class"])))
comp.wide <- ungroup(comp.wide) %>%  select(one_of(waps.specs$Newval[waps.specs$Type != "class"]))
comp.wide <- apply(comp.wide, 2, as.numeric)
comp.wide[is.na(comp.wide)] <- 0

# Merges with treatment labels
comp.attr <- left_join(comp.attr, trtlabels)
comp.attr$sp.trt <- gsub("[[:space:]]", "", comp.attr$sp.trt)
```