---
title: "Classification - Part 2"
output:
  pdf_document: default
  html_document:
    df_print: paged
  highlight: github
  prettydoc::html_pretty: null
---


# Fitting Multi-state (Markov) models with MSM

Multi-state Markov models might be a relevant way to analyze this data, where
repeated observations of a single subject are quantified as a set of discrete
states with temporal transitions in state assignment described by a matrix
of transition properties. Also possible with these models is the addition of
covariates that may change transition probabilities, rather than operating with
only a fixed probability for each given state change.

```{r, echo = FALSE, warning = FALSE, message=FALSE}
library(lubridate); library(tidyverse); library(msm); library(tidyquant)

clustdat <- read.csv("../data/WAPSclusters.csv",
                     header = TRUE,
                     stringsAsFactors = FALSE)

climate <- read.csv("../data/cimis_monthly.csv",
                       header = TRUE,
                       stringsAsFactors = FALSE)

climate_daily <- read.csv("../data/cimis_daily.csv",
                       header = TRUE,
                       stringsAsFactors = FALSE)


labellist <- c('1. Native Perennials',
                    '2. Festuca - B. hordeaceous Annuals',
                    '3. Invasive Annuals',
                    '4. Avena - B. diandrus Annuals')
```

# Raw matrix of transition probabilities

The simplest way to assess differences in transition probability is to just plot
the total number of transition events observed within the dataset. This matrix,
shown below, describes the general pattern visualized earlier in the plotting of
states over time:

* Both the "natives" and "WAPS" states are relatively stable over the time period
we observed.

* All states were quite likely to shift to the "dry annuals" state (likely the
result of the influx of Avena / Bromus diandrus during the peak of the drought)

* "Priority Annuals", which did very well once initially planted, were quick to be lost
from the dataset.

```{r, echo = FALSE, warning = FALSE}
st = as.matrix(statetable.msm(cluster, plot, data = clustdat %>% arrange(plot, year)))
colnames(st) = c("Natives", "WetA", "WAPS", "DryA")
rownames(st) = c("Natives", "WetA", "WAPS", "DryA")
st
```

# Fitting state-transition models:

Fitting these models is relatively simply using the MSM package -- the function
attempts to find the best values of the transition matrix that predict the observed
frequencies of each state over time.

MSMs also allow for the use of likelihood ratio tests -- quantitative measures
for how additional covariates in the model improve fit. However, one important
caveat is that each element of the transition matrix is fit by a unique parameter
(12 total for a 4x4 matrix, as diagonal entries are equal to 1 - sum of all
transition probabilities), so the addition of a covariate must change a number of
unique parameters, greatly reducing the residual degrees of freedom.

```{r, echo = FALSE, warning = FALSE}
# Dummy vectors for priority effects
clustdat$waps = as.numeric(grepl("WAPS", clustdat$sp.trt))
clustdat$natives = as.numeric(grepl("natives", clustdat$sp.trt))
clustdat$annuals = as.numeric(grepl("annuals", clustdat$sp.trt))

# Defining initial matrix
Q <- rbind ( c(0, 0.333, 0.333, 0.333),
             c(0.333, 0, 0.333, 0.333),
             c(0.333, 0.333, 0, 0.333),
             c(0.333, 0.333, 0.333, 0))

Q.crude <- crudeinits.msm(cluster ~ year, plot, data=clustdat, qmatrix=Q)

# Base model
msm.null <- msm(cluster ~ year, plot, data=clustdat, qmatrix=Q, obstype=2)
```

To assess whether priority effects impact transition probabilities, we can include
planting mixture effects on transition probabilities. To reduce the total number
of degrees of freedom associated with each covariate, I've restricted their
impact to only those transitions from one state to a group found in the initial
planting composition. E.g. asking whether the addition of natives to the initial
planting group influences the likelihood that a given community will later
transition to a native state. This would only affect 3 elements of the transition
matrix governing the transitions to state 1 (native state) from states
2, 3, or 4 (wet annual, WAPS, or dry annual states, respectively).

The resulting likelihood ratio test suggests that the addition of these parameters
significantly improves model fit (p > .001).

```{r, echo = FALSE, warning = FALSE}
# With priority effects
msm.priority <- msm(cluster ~ year, plot, data=clustdat, qmatrix=Q, obstype=2,
                    covariates = list("1-2" = ~ annuals,
                                      "1-3" = ~ waps,
                                      "1-4" = ~ annuals,
                                      "2-1" = ~ natives,
                                      "2-3" = ~ waps,
                                      "2-4" = ~ annuals,
                                      "3-1" = ~ natives,
                                      "3-2" = ~ annuals,
                                      "3-4" = ~ annuals,
                                      "4-1" = ~ natives,
                                      "4-2" = ~ annuals,
                                      "4-3" = ~ waps))
cat("Likelihood ratio test: \n")
lrtest.msm(msm.null, msm.priority)
```

We can also visualize how well our model explains the observed patterns using
a simple prevalence plot, which shows expected frequencies of a state in red
with estimated state frequencies in blue.  

I think that this model does quite a good job for two of the different states --
Natives (state 1), and Priority Annuals (state 2), which are either quite stable over time,
or decrease rapidly.  

However, we seem to do poorly with WAPS and Dry Annuals states, whose transition
probabilities seem to correlate strongly with climatic variation.

```{r, echo = FALSE, warning = FALSE}
plot.prevalence.msm(msm.priority, mintime=2008, maxtime=2018)
cat("1 - Native, 2 - Priority Annuals, 3 - WAPS, 4 - Dry Annuals")
```

# Including climate data into analyses:

To improve the model fit above, we could try to incorporate environmental covariates
into our data as well. We know that precipitation patterns can have major effects
on community assembly, but it is unclear how to best examine the influence of
this driver on state change rates.  

To start, figures below show data on monthly precipitation and evapotranspiration
measures taken from the Davis CIMIS (irrigation management) system. These fluctuations
show a somewhat expected pattern - years 2012 - 2015 show a profound drought,
while the periods between 2008 - 2011 and 2015-2017 are quite wet.

```{r, echo = FALSE, warning = FALSE}
climate_daily$Date = as.POSIXct(climate_daily$Date, format = "%m/%d/%Y")

climate_daily %>%
  group_by(month=floor_date(Date, "month")) %>%
  summarise(precip = sum(na.omit(Precip..mm.)),
            eto = sum(na.omit(ETo..mm.))) %>%
  ggplot(aes(x = month, y = precip)) +
  geom_line() +
  xlab("Date (Aggregated by Month)") +
  ylab("Total Precip (mm)") +
  ggtitle("Precipitation by Month 2006 - 2018")

climate_daily %>%
  group_by(month=floor_date(Date, "month")) %>%
  summarise(precip = sum(na.omit(Precip..mm.)),
            eto = sum(na.omit(ETo..mm.))) %>%
  ggplot(aes(x = month, y = precip - eto)) +
  geom_line()+
  xlab("Date (Aggregated by Month)") +
  ylab("Precip - ETo (mm)") +
  ggtitle("Net Water Deficit by Month 2006 - 2018")
```

Many studies focusing on climate-driven changes in species abundance have focused on
early-season precipitation as a main determinant of community composition. As a start,
I will focus here on precipitation and ETo values between December and February.

There are a number of different ways to use this data to evluate changes in
community transition rates -- total precipitation, net water availability
(precip - ETo), and lagged precipitation, to name a few. Dudney et al (2017),
which attempts to evaluate current year and lagged precipitation effects, may
be of some use.  

The figure below shows this early precipitation change, aggregated into monthly
intervals, with a three-year moving average indicated by the dashed line.

```{r, echo = FALSE, warning = FALSE}
climate_daily %>% mutate(month = month(Date),
                         year = year(Date)) %>%
  group_by(Water.Year) %>%
  filter(month %in% c(11,12,1,2)) %>%
  summarise(precip = sum(na.omit(Precip..mm.)),
            eto = sum(ETo..mm.)) %>%
  ggplot() +
  geom_bar(stat = "identity",
           aes(y = precip,
               x = Water.Year,
               fill = "blue"),
           width = .3, color = "black") +
    geom_bar(stat = "identity",
           aes(y = eto,
               x = Water.Year + .3,
               fill = "tan"),
           width = .3, color = "black") +
    geom_bar(stat = "identity",
           aes(y = precip - eto,
               x = Water.Year + .6,
               fill = "green"),
           width = .3, color = "black") +
  geom_ma(aes(y = precip - eto,
              x = Water.Year + .6),
              color = "black", n = 3) +
    scale_fill_identity(name = 'Measure', guide = 'legend', labels = c('Precip (mm)', 'ETo (mm)', 'Net (mm)')) +
  xlab("Year") +
  ylab("Water Availability (Nov - Feb") +
  ggtitle("CIMIS Early Season Water Availability")
```

Because I don't seem to see a consistent trend here where lower values of net
precipitation minus ETo produce within year-effects, I think a running average
approach might be best. Rather than accounting for just a single year of data,
taking a running average of drought stress can account for successive effects
of multiple years.

```{r, echo = FALSE, warning = FALSE, message=FALSE}
climate$month = gsub("-.*", "", climate$Month.Year)

climsum = climate_daily %>% mutate(month = month(Date), year = year(Date)) %>%
  group_by(Water.Year) %>%
  filter(month %in% c(11,12,1,2)) %>%
  summarise(precip = sum(na.omit(Precip..mm.)),
            eto = sum(ETo..mm.),
            net = precip - eto) %>%
  mutate(threeyear_ma = SMA(net, 3),
         net_cum_2 = c(NA, rollsum(net, 2)),
         net_cum_3 = c(NA, NA, rollsum(net, 3)))

# Adding weather data
clustdat = left_join(clustdat, climsum %>% rename(year = Water.Year))

# With environmental covariates
msm.env <- msm(cluster ~ year, plot, data=clustdat, qmatrix=Q, obstype=2, covariates = ~ net_cum_3)

msm.env_priority <- msm(cluster ~ year, plot, data=clustdat, qmatrix=Q, obstype=2,
                    covariates = list("1-2" = ~ annuals + net_cum_3,
                                      "1-3" = ~ waps + net_cum_3,
                                      "1-4" = ~ annuals + net_cum_3,
                                      "2-1" = ~ natives + net_cum_3,
                                      "2-3" = ~ waps + net_cum_3,
                                      "2-4" = ~ annuals + net_cum_3,
                                      "3-1" = ~ natives + net_cum_3,
                                      "3-2" = ~ annuals + net_cum_3,
                                      "3-4" = ~ annuals + net_cum_3,
                                      "4-1" = ~ natives + net_cum_3,
                                      "4-2" = ~ annuals + net_cum_3,
                                      "4-3" = ~ waps + net_cum_3))
cat("Likelihood ratio test (with env. covariates vs. null): \n")
lrtest.msm(msm.null, msm.env)
cat("Likelihood ratio test (with env. covariates + priority vs. priority: \n")
lrtest.msm(msm.priority, msm.env)
cat("AIC scores of all models (lowest score is best): \n")
data.frame(model = c("Null", "Priority", "Environment", "Env + Priority"),
           AIC = c(AIC(msm.null), AIC(msm.priority), AIC(msm.env), AIC(msm.env_priority)))
```

The subsequent prevalence plot shows a substantially better fit to the models with
this environmental data included - though it's not perfect.

```{r, echo = FALSE, warning = FALSE}
# Estimated values of Q
qmat = as.matrix(qmatrix.msm(msm.env_priority)$estimates)
colnames(qmat) = c("Natives", "WetA", "WAPS", "DryA")
rownames(qmat) = c("Natives", "WetA", "WAPS", "DryA")

# With both
# plot.prevalence.msm(msm.l1_priority, mintime=2008, maxtime=2018)
# cat("1 - Native, 2 - Priority Annuals, 3 - WAPS, 4 - Dry Annuals")
```

# Next steps:

Going forward, I think the most important ways to refine this analysis are:

1. Make clustering approaches more concrete - what are the best steps in subsetting
data and classifying community types? $\checkmark$

2. Define questions clearly - what exactly is this approach answering? $\checkmark$?

3. Determine the best way to analyze climatic data - how to include variables
in transition models? $\checkmark$

4. Follow-up with clearer visualizations of how climate impacts persistence / state change -
visualize how probabilities of transitions change with priority effects across a
gradient of climatic values. $\checkmark$

 5. Identify other tests and analyses, as needed.

# Best way to analyze climatic data:

I need to double-check how the model functions -- are transition probabilities a
function of a water year's precipitation (i.e. are we predicting transitions in
year t + 1 using data at year t? I think this is likely the case) or as a function
of the next year's precipitation? The model that best corresponds to our conceptual
notion of state change isn't climate at t, but climate at t + 1 (it's between our
two sets of observations, but impacts the next year).

The best response, based on other publications is to quantify drought stress using
SPEI, a simple, standardized metric that is broadly applicable. See "ExploringClimateData"
for more information on its calculation.

```{r}
speidata <- read.csv("../data/SPEIfits.csv")
colnames(speidata)
clustdat_t1 <- clustdat %>%
  left_join((speidata %>% mutate(year = year + 1)), by = c("year" = "year"))

msm.l1_priority <- msm(cluster ~ year, plot, data=clustdat_t1, qmatrix=Q, obstype=2,
                    covariates = list("1-2" = ~ annuals + l1fit,
                                      "1-3" = ~ waps + l1fit,
                                      "1-4" = ~ annuals + l1fit,
                                      "2-1" = ~ natives + l1fit,
                                      "2-3" = ~ waps + l1fit,
                                      "2-4" = ~ annuals + l1fit,
                                      "3-1" = ~ natives + l1fit,
                                      "3-2" = ~ annuals + l1fit,
                                      "3-4" = ~ annuals + l1fit,
                                      "4-1" = ~ natives + l1fit,
                                      "4-2" = ~ annuals + l1fit,
                                      "4-3" = ~ waps + l1fit))

msm.l2_priority <- msm(cluster ~ year, plot, data=clustdat_t1, qmatrix=Q, obstype=2,
                    covariates = list("1-2" = ~ annuals + l2fit,
                                      "1-3" = ~ waps + l2fit,
                                      "1-4" = ~ annuals + l2fit,
                                      "2-1" = ~ natives + l2fit,
                                      "2-3" = ~ waps + l2fit,
                                      "2-4" = ~ annuals + l2fit,
                                      "3-1" = ~ natives + l2fit,
                                      "3-2" = ~ annuals + l2fit,
                                      "3-4" = ~ annuals + l2fit,
                                      "4-1" = ~ natives + l2fit,
                                      "4-2" = ~ annuals + l2fit,
                                      "4-3" = ~ waps + l2fit))

msm.l3_priority <- msm(cluster ~ year, plot, data=clustdat_t1, qmatrix=Q, obstype=2,
                    covariates = list("1-2" = ~ annuals + l3fit,
                                      "1-3" = ~ waps + l3fit,
                                      "1-4" = ~ annuals + l3fit,
                                      "2-1" = ~ natives + l3fit,
                                      "2-3" = ~ waps + l3fit,
                                      "2-4" = ~ annuals + l3fit,
                                      "3-1" = ~ natives + l3fit,
                                      "3-2" = ~ annuals + l3fit,
                                      "3-4" = ~ annuals + l3fit,
                                      "4-1" = ~ natives + l3fit,
                                      "4-2" = ~ annuals + l3fit,
                                      "4-3" = ~ waps + l3fit))
cat("T1 Model")
AIC(msm.env_priority, msm.l1_priority, msm.l2_priority, msm.l3_priority)

```

T1 approach seems to be better via AIC, but model doesn't appear to work right in reality.

```{r}
clustdat_t0 <- clustdat %>%
   left_join((speidata %>% mutate(year = year)), by = c("year" = "year"))

 msm.l1_priority <- msm(cluster ~ year, plot, data=clustdat_t0, qmatrix=Q, obstype=2,
                     covariates = list("1-2" = ~ annuals + l1fit,
                                       "1-3" = ~ waps + l1fit,
                                       "1-4" = ~ annuals + l1fit,
                                       "2-1" = ~ natives + l1fit,
                                       "2-3" = ~ waps + l1fit,
                                       "2-4" = ~ annuals + l1fit,
                                       "3-1" = ~ natives + l1fit,
                                       "3-2" = ~ annuals + l1fit,
                                       "3-4" = ~ annuals + l1fit,
                                       "4-1" = ~ natives + l1fit,
                                       "4-2" = ~ annuals + l1fit,
                                       "4-3" = ~ waps + l1fit))

 msm.l2_priority <- msm(cluster ~ year, plot, data=clustdat_t0, qmatrix=Q, obstype=2,
                     covariates = list("1-2" = ~ annuals + l2fit,
                                       "1-3" = ~ waps + l2fit,
                                       "1-4" = ~ annuals + l2fit,
                                       "2-1" = ~ natives + l2fit,
                                       "2-3" = ~ waps + l2fit,
                                       "2-4" = ~ annuals + l2fit,
                                       "3-1" = ~ natives + l2fit,
                                       "3-2" = ~ annuals + l2fit,
                                       "3-4" = ~ annuals + l2fit,
                                       "4-1" = ~ natives + l2fit,
                                       "4-2" = ~ annuals + l2fit,
                                       "4-3" = ~ waps + l2fit))

 msm.l3_priority <- msm(cluster ~ year, plot, data=clustdat_t0, qmatrix=Q, obstype=2,
                     covariates = list("1-2" = ~ annuals + l3fit,
                                       "1-3" = ~ waps + l3fit,
                                       "1-4" = ~ annuals + l3fit,
                                       "2-1" = ~ natives + l3fit,
                                       "2-3" = ~ waps + l3fit,
                                       "2-4" = ~ annuals + l3fit,
                                       "3-1" = ~ natives + l3fit,
                                       "3-2" = ~ annuals + l3fit,
                                       "3-4" = ~ annuals + l3fit,
                                       "4-1" = ~ natives + l3fit,
                                       "4-2" = ~ annuals + l3fit,
                                       "4-3" = ~ waps + l3fit))

 cat("T0 Model")
 AIC(msm.l1_priority, msm.l2_priority, msm.l3_priority)
```

# Goodness-of-fit test

```{r}
outputtable = data.frame(obs = colSums(prevalence.msm(msm.l1_priority)$Observed),
exp = colSums(prevalence.msm(msm.l1_priority)$Expected))

chisq.test(outputtable$obs[c(1:4)], outputtable$exp[c(1:4)])
```

# Bootstrapping confidence intervals (too computationally intensive to be useful)

```{r}
# system.time(qmatrix.msm(msm.l1_priority, covariates=list(l1fit = netprecip,
#                                              natives = pri,
#                                              annuals = pri,
#                                              waps = pri),
#             ci = "boot", B = 10))
# 
# system.time(qmatrix.msm(msm.l1_priority, covariates=list(l1fit = netprecip,
#                                              natives = pri,
#                                              annuals = pri,
#                                              waps = pri),
#             ci = "boot", B = 100))
# 
# 
# system.time(qmatrix.msm(msm.l1_priority, covariates=list(l1fit = netprecip,
#                                              natives = pri,
#                                              annuals = pri,
#                                              waps = pri),
#             ci = "boot", B = 1000))
```


# Visualizing response to climate:

Look at raw transition probabilities (add covariates list)

See MSM documentation on how to estimate probabilities with covariates,
stability, etc.

Formulation of transition probabilities from Qmatrix:

When all transitions are possible, the probability of transitioning to either states 2, 3, or 4 from state 1 can be expressed as:

$$ P_{11} = e^{-(q_{12} + q_{13} + q_{14})} $$

and

$$ P_{12} = 1 - e^{-q_{12}} $$
$$ P_{13} = 1 - e^{-q_{13}} $$
$$ P_{14} = 1 - e^{-q_{14}} $$

```{r}
qmatrix.msm(msm.l1_priority, covariates=list(l1fit = 0))
p12 = 1 - exp(-.02581)
p13 = 1 - exp(-.02605)
p14 = 1 - exp(-.1894)
tot = sum(p12,p13,p14)
p12/tot
p13/tot


msm.l1_priority <- msm(cluster ~ year, subject = plot, data=clustdat_t0, qmatrix=Q, obstype=2,
                    covariates = list("1-2" = ~ annuals + l1fit,
                                      "1-3" = ~ waps + l1fit,
                                      "1-4" = ~ annuals + l1fit,
                                      "2-1" = ~ natives + l1fit,
                                      "2-3" = ~ waps + l1fit,
                                      "2-4" = ~ annuals + l1fit,
                                      "3-1" = ~ natives + l1fit,
                                      "3-2" = ~ annuals + l1fit,
                                      "3-4" = ~ annuals + l1fit,
                                      "4-1" = ~ natives + l1fit,
                                      "4-2" = ~ annuals + l1fit,
                                      "4-3" = ~ waps + l1fit),
                    exacttimes = TRUE)
haz = hazard.msm(msm.l1_priority)
hazard.msm(msm.l1_priority, cl = .99)
?hazard.msm
```

```{r}
for(i in 1:length(haz)){

  newdf = data.frame(haz[[i]])
  newdf$coef = names(haz)[i]
  newdf$trans = rownames(newdf)

  if(i == 1){
    outputdf = newdf
  }else{
    outputdf = bind_rows(list(outputdf, newdf))
  }

}

hazard_rats = outputdf %>% filter(HR != 1 & U != 1 & L != 1)
write.csv(x = hazard_rats, "../data/hazardrates.csv")
```

Below is code to pull out transition probabilities for each state.
However, there is likely a better way:

* Create matrix of Q values
* Create function that extracts probabilities based on the function exp(q) or
1 - exp(-q).
* Assign diagonals to the probability of maintaining a state
* Assign non-diagonals to the probability of transitioning
* Repeat for confidence intervals (both low and high)

```{r}
Nprob = list()
counter = 1

for(pri in c(0:1)){
  for(netprecip in seq(-1.5, 1.5, by = .1)){
    qm = as.matrix(qmatrix.msm(msm.l1_priority, covariates=list(l1fit = netprecip,
                                                               natives = pri, annuals = pri, waps = pri))$estimates)
    ql = as.matrix(qmatrix.msm(msm.l1_priority, covariates=list(l1fit = netprecip,
                                                                 natives = pri, annuals = pri, waps = pri))$L)
    qu = as.matrix(qmatrix.msm(msm.l1_priority, covariates=list(l1fit = netprecip,
                                                                 natives = pri, annuals = pri, waps = pri))$U)

    qdf = data.frame(qm) %>% mutate(From = rownames(qm)) %>%
      gather("key" = "To", "value" = "Trans", -From) %>%
      mutate(To = gsub("\\.", " ", To))

    qlow = data.frame(ql) %>% mutate(From = rownames(ql)) %>%
      gather("key" = "To", "value" = "Lower", -From) %>%
      mutate(To = gsub("\\.", " ", To))

    qupp = data.frame(qu) %>% mutate(From = rownames(qu)) %>%
      gather("key" = "To", "value" = "Upper", -From) %>%
      mutate(To = gsub("\\.", " ", To))

    qdf$Trans = if_else(qdf$From == qdf$To, exp(qdf$Trans), 1 - exp(-qdf$Trans))
    qlow$Lower = if_else(qlow$From == qlow$To, exp(qlow$Lower), 1 - exp(-qlow$Lower))
    qupp$Upper = if_else(qupp$From == qupp$To, exp(qupp$Upper), 1 - exp(-qupp$Upper))

    output <- left_join(qdf, qlow) %>% left_join(qupp)
    output$precip = rep(netprecip, nrow(output))
    output$priority = rep(pri, nrow(output))

    Nprob[[counter]] <- output
    counter = counter + 1
  }
}


cols = rainbow(4, v = .5)

bind_rows(Nprob) %>%
  filter(From == To) %>%
  mutate(priority = if_else(priority == 1, "In Mixture", "Alone")) %>%
  ggplot(aes(x = precip,
             color = From,
             fill = From)) +
  geom_line(aes(y = Trans), size = 1.5) +
  theme_bw() +
  #geom_ribbon(aes(ymax = Upper, ymin = Lower), alpha = .2) +
  facet_wrap(~priority) +
  ggtitle("Climate and Seeding Composition Effects on State Resilience") +
  scale_color_manual(name = 'State Assignment',
         values =c("State 1" = cols[1],
                   "State 2" = cols[2],
                   "State 3" = cols[3],
                   "State 4" = cols[4]),
         labels = labellist) +

  scale_fill_manual(name = 'State Assignment',
         values =c("State 1" = cols[1], "State 2" = cols[2],
                   "State 3" = cols[3], "State 4" = cols[4]),
         labels = labellist) +

  xlab("SPEI") +
  ylab("Probability of Retaining State in Next Year")

# ggsave("../figures/StabilityProbabilities_NoErrorBars.pdf", height = 4, width = 7)

```

```{r}
trtnames = c("to Natives", "to Priority Annuals", "to WAPS", "to Dry Annuals")
priorities = c("No Priority", "Priority")

tprob = function(stateval = "State 1", title = "E. glaucus - S. pulchra"){

  p = bind_rows(Nprob)

  p$Trans[p$From == stateval & p$To == stateval] = -1

  plot = p %>%
  # filter(From != To) %>%
  filter(To == stateval) %>%
  mutate(priority = if_else(priority == 1, "Represented In Seeding", "Not Represented")) %>%
  ggplot(aes(x = precip,
             color = From,
             fill = From)) +
  geom_line(aes(y = Trans), size = 1.5) +
  theme_bw() +
  #geom_ribbon(aes(ymax = Upper, ymin = Lower), alpha = .2) +
  facet_grid(~priority) +
  ggtitle(title) +
  scale_color_manual(name = 'Previous State Assignment',
         values =c("State 1" = cols[1], "State 2" = cols[2],
                   "State 3" = cols[3], "State 4" = cols[4]),
         labels = labellist) +

  scale_fill_manual(name = 'Initial State',
         values =c("State 1" = cols[1], "State 2" = cols[2],
                   "State 3" = cols[3], "State 4" = cols[4]),
         labels = labellist) +

  xlab("SPEI") +
  ylab("Probability of Assignment in Next Year") +
  ylim(0, .5)

  return(plot)
}

alltogether = gridExtra::grid.arrange(tprob(),
                        tprob("State 2", labellist[2]),
                        tprob("State 3", labellist[3]),
                        tprob("State 4", labellist[4]),
                        nrow = 4,
                        ncol = 1,
                        top = "Probability of Transition to States")
 
# ggsave(plot = alltogether,
#        "../figures/TransitionProbabilities_NoErrorBars_v2.pdf", height = 14, width = 8)
```

# Highlighting particular transitions

## Transition to native-dominated state

```{r}
stateval = "State 1"
title = "E. glaucus - S. pulchra"
textdf = data.frame(
      precip = c(0,0),
      Trans = c(.5, .5),
      lab = c("+ Priority", "- Priority"),
      priority = c("E. glaucus - S. pulchra ", "E. glaucus - S. pulchra"))
p = bind_rows(Nprob)

  # p$Trans[p$From == stateval & p$To == stateval] = -1

native_trans = p %>%
  filter(From != To) %>%
  filter(To == stateval) %>%
  mutate(To = plyr::revalue(To,
                            c("State 1" = 'E. glaucus - S. pulchra'))) %>%
  mutate(priority = if_else(priority == 1, "+ Priority ", "- Priority")) %>%
  ggplot(aes(x = precip,
             color = From,
             fill = From)) +
  geom_line(aes(y = Trans, linetype = From), size = 1.5) +
  theme_bw() +
  #geom_ribbon(aes(ymax = Upper, ymin = Lower), alpha = .2) +
  facet_grid(~priority) +
  ggtitle("Priority Affects Native Perennial Transition Probability") +
  scale_color_manual(name = 'Previous State Assignment',
                     values =c("State 2" = cols[2],
                               "State 3" = cols[3], "State 4" = cols[4]),
                     labels = labellist[2:4]) +

  scale_linetype_manual(name = 'Previous State Assignment',
                        values =c("State 2" = 2,
                                  "State 3" = 1, "State 4" = 1),
                     labels = labellist[2:4]) +

  xlab("SPEI") +
  ylab("Native Perennial Transition Probability") +
  ylim(0, .5)
  # geom_text(data = textdf, aes(fill = NULL, color = NULL, y= Trans), label = textdf$lab, show.legend = FALSE)

native_trans
ggsave("../figures/eglaspul_transition.pdf", width = 7, height = 4)
```

## Transitions between invasive and drought tolerant states

```{r}
stateval = c("State 3", "State 4")

p = bind_rows(Nprob)

  # p$Trans[p$From == stateval & p$To == stateval] = -1
ann_change = p %>%
  group_by(From, To, precip) %>%
  filter(priority == 0) %>%
  filter(From != To) %>%
  filter(From %in% stateval) %>%
  mutate(pval = case_when(
    From == "State 3" & To %in% c("State 1", "State 4") ~ .05,
    From == "State 4" & To == "State 1" ~ .05,
    TRUE ~ 1)) %>%
  ungroup() %>%
  # mutate(To = plyr::revalue(To,
  #                           c("State 1" = 'E. glaucus - S. pulchra',
  #                             "State 2" = 'F. perennis - B. hordeaceous',
  #                             "State 3" = 'E. caput-medusae - A. triuncialis',
  #                             "State 4" = 'A. fatua - B. diandrus'))) %>%
  # mutate(priority = if_else(priority == 1, "Represented In Seeding", "Not Represented")) %>%
  ggplot(aes(x = precip)) +
  geom_line(aes(y = Trans, color = To, linetype = as.factor(pval)), size = 1.5) +
  theme_bw() +
  #geom_ribbon(aes(ymax = Upper, ymin = Lower), alpha = .2) +
  facet_grid(~From) +
  ggtitle("B) Drought Stress Effects on Transition Probability") +
  scale_color_manual(name = 'Subsequent State Assignment',
                     values =c("State 1" = cols[1], "State 2" = cols[2],
                               "State 3" = cols[3], "State 4" = cols[4]),
                     labels = labellist) +

  xlab("SPEI") +
  ylab("Transition Probability") +
  ylim(0, .2) +
  guides(linetype = FALSE)

coef_highlight = gridExtra::grid.arrange(native_trans, ann_change, nrow = 2)
ggsave(plot = coef_highlight, "../figures/coefficient_highlight.pdf", width = 6, height = 6)
```
