---
title: "Priority Effect Analysis"
author: "Evan Batzer"
date: "August 20, 2018"
output: html_document
---

# State change ideas:

* What are relevant groupings of compositional states within the community?
  * Does the native/waps/annuals/weeds distinction hold overall (does it explain much of community variance?)
  
* What is the initial state of communities based on planting treatment?
  * When planting multiple groups, which one wins?
  
* How long is that initial state retained (community stability)?

* If transition occurs, from what to state to what state?
  * Does it couple with precipitation?
  * Does species planting matter too?
  
* When transition occurs, what characteristics are associated with the transition?
  * Nutrient availability in year - 1
  * Precipitation in currrent and previous year
  * Biomass in the previous year
  
* If precipitation matters, when in the year does it fall?
  * Fall vs. winter vs. spring precip


# Similarity comparisons
## How compositionally different are the different species plantings to one another over time?

Do we see convergence, divergence, or no consistent pattern in dissimilarity among treatment groups?


```{r}
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)
library(vegan)

# Cover data (wide format)
cover <- read.csv("../data/comdata_wideformat.csv", 
                  header = TRUE, 
                  stringsAsFactors = FALSE)
group_percents = cover[,12:23]
cover = cover[,-(12:23)]

# Treatment labels and group structure
trtlabels <- read.csv("../data/trtlabels.csv",
          header = TRUE,
          stringsAsFactors = FALSE)

# Biomass data
biomass <-  read.csv("../data/biomass_all.csv", 
                  header = TRUE, 
                  stringsAsFactors = FALSE)

# Other treatment labels (checking for redundancy)
checklabels <- read.csv("../data/plot_ids.csv",
          header = TRUE,
          stringsAsFactors = FALSE)

checklabels$Sp.comp <- gsub(" ", "", checklabels$Sp.comp)

# Species treatments
waps.specs <- read.csv("../data/WAPSspecies.csv",
                       header = TRUE,
                       stringsAsFactors = FALSE)

# Functional group percent composition
funcgroups <- read.csv("../data/func_group_percents.csv",
                       header = TRUE,
                       stringsAsFactors = FALSE)

funcgroups = funcgroups %>% select(-X)
```

# Analyzing Planting Group (WAPS, annuals, natives) Change over Time

```{r}
# Assign functional group composition dataset
pe.com = funcgroups 

# Standardize to relative cover
pe.com[,4:ncol(pe.com)] = vegan::decostand(pe.com[,4:ncol(pe.com)], method = "total")

# Convert to long format for plotting
pe.com_long = gather(pe.com, key = "class", value = "cover", -c(plot, year, sp.trt))

# Join with treatment labels
pe.com_long = left_join(pe.com_long, trtlabels, by = "plot") %>% 
  filter(fert == "none" & clip == "none")

# Define plotting function
plot_fgs <- function(water){
  
  p = pe.com_long %>% 
  filter(h2o == water) %>% # Filter by h2o treatment (water) name
  ggplot(aes(x = year,
             y = cover,
             color = class)) +
  geom_point() + 
  stat_smooth(se = FALSE) + 
  ylim(0,1.1) +
  ggtitle(paste(water, "H2O")) +
  facet_wrap(~sp.trt.x) +
  scale_x_continuous(breaks = seq(2008,2018, by = 2), 
                   labels = as.character(seq(2008,2018, by = 2))) +
  ylab("Proportion Total Cover")
  
  return(p)
}

# Control treatment
plot_fgs("control")

# Dry treatment
plot_fgs("dry")

# Wet treatment
plot_fgs("wet")

# Late wet treatment
plot_fgs("late wet")

# Control 2nds
plot_fgs("control-2nd")

```

# Dissimilarity within blocks

```{r}
selected.trts <- c("annuals", "WAPS", "Natives", "natives+WAPS+annuals", 
                   "annuals+natives", "WAPS+annuals", "WAPS+natives")

cover = cover %>% select(-X) %>% filter(sp.trt %in% selected.trts)
cover[,1:10] =  lapply(cover[, 1:10], as.factor)
rs = cover %>% select_if(is.numeric) %>% do(data.frame(rowSums(.)))
cover = cover %>% filter(rs != 0)

com.mat = cover %>% select_if(is.numeric)
com.attr = cover %>% select_if(is.factor)

# braydisp = betadisper(vegdist(com.mat, method = "bray"), 
#                       paste(com.attr$year, com.attr$block, com.attr$subblock, sep = "_"), sqrt.dist = TRUE)
# 
# eucdisp = betadisper(vegdist(com.mat, method = "euclidean"),
#                      paste(com.attr$year, com.attr$block, com.attr$subblock, sep = "_"), sqrt.dist = FALSE)
# 
# bd = data.frame(grp = braydisp$group, bray = braydisp$distances, euc = eucdisp$distances)
# 
# grpdisp = bind_cols(com.attr, 
#                     data.frame(grp = bd$grp, bray = bd$bray, euc = bd$euc))

```

# Plotting dispersion as a function of water treatment:


```{r}
# library(gridExtra)
# 
# # Define plotting function
# plot_fgs <- function(water){
#   
#   p = grpdisp %>% 
#     filter(clip == "none" & fertilization == "none") %>%
#     filter(h2o == water) %>% # Filter by h2o treatment (water) name
#     ggplot(aes(x = as.numeric(year),
#                y = bray ^ 2,
#                color = sp.trt)) +
#     geom_point() + 
#     stat_smooth(se = FALSE) + 
#     ylim(0,1) +
#     ggtitle(paste(water, "H2O")) +
#     scale_x_continuous(breaks = seq(0,10, by = 2), 
#                      labels = as.character(seq(2008,2018, by = 2))) +
#     ylab("Distance to Block Centroid (Bray-Curtis)") +
#     xlab("Year") +
#     stat_smooth(aes(color = NULL), color = 'black')
#   
#   return(p)
# }
# 
# for(i in unique(grpdisp$h2o)){
#   p = plot_fgs(i)  
#   plot(p)
# }
```

# Plotting dispersion as a function of treatment:

```{r}
# Define plotting function
# plot_fgs <- function(sptrt){
#   
#   p = grpdisp %>% 
#     filter(h2o != "control-2nd") %>%
#     filter(clip == "none" & fertilization == "none") %>%
#     filter(sp.trt == sptrt) %>% # Filter by h2o treatment (water) name
#     ggplot(aes(x = as.numeric(year),
#                y = bray ^ 2,
#                color = h2o)) +
#     geom_point() + 
#     stat_smooth(se = FALSE) + 
#     ylim(0,1) +
#     ggtitle(paste(sptrt, "Planting Comp")) +
#     scale_x_continuous(breaks = seq(0, 10, by = 2), 
#                      labels = as.character(seq(2008,2018, by = 2))) +
#     ylab("Distance to Block Centroid (Bray-Curtis)") +
#     xlab("Year")
#   
#   return(p)
# }
# 
# for(i in unique(grpdisp$sp.trt)){
#   p = plot_fgs(i)  
#   plot(p)
# }

```

# How dissimilar are groups of the same planting treatment over time:

```{r}
# selected.trts <- c("annuals", "WAPS", "Natives", "natives+WAPS+annuals", 
#                    "annuals+natives", "WAPS+annuals", "WAPS+natives")
# 
# # cover = cover %>% filter(sp.trt %in% selected.trts)
# # cover[,1:10] =  lapply(cover[, 1:10], as.factor)
# # rs = cover %>% select_if(is.numeric) %>% do(data.frame(rowSums(.)))
# # cover = cover %>% filter(rs != 0)
# 
# com.mat = cover %>% select_if(is.numeric)
# com.attr = cover %>% select_if(is.factor)
# 
# braydisp = betadisper(vegdist(com.mat, method = "bray"), 
#                       paste(com.attr$year, com.attr$h2o, com.attr$sp.trt, 
#                             com.attr$clip, com.attr$fertilization, sep = "_"), sqrt.dist = TRUE)
# 
# eucdisp = betadisper(vegdist(com.mat, method = "euclidean"),
#                       paste(com.attr$year, com.attr$h2o, com.attr$sp.trt, 
#                             com.attr$clip, com.attr$fertilization, sep = "_"), sqrt.dist = FALSE)
# 
# bd = data.frame(grp = braydisp$group, bray = braydisp$distances, euc = eucdisp$distances)
# 
# grpdisp = bind_cols(com.attr, 
#                     data.frame(grp = bd$grp, bray = bd$bray, euc = bd$euc))
# 
# # Define plotting function
# plot_fgs <- function(water){
#   
#   p = grpdisp %>% 
#     filter(h2o == water) %>%
#     filter(clip == "none" & fertilization == "none") %>%
#     ggplot(aes(x = as.numeric(year),
#                y = bray ^ 2,
#                color = sp.trt)) +
#     geom_point() + 
#     stat_smooth(se = FALSE) +
#     ylim(0,1) +
#     ggtitle(paste(water, "H2O")) +
#     scale_x_continuous(breaks = seq(0, 10, by = 2), 
#                      labels = as.character(seq(2008,2018, by = 2))) +
#     ylab("Distance to Block Centroid (Bray-Curtis)") +
#     xlab("Year")
#   
#   return(p)
# }
# 
# for(i in unique(grpdisp$h2o)){
#   p = plot_fgs(i)  
#   plot(p)
# }

```


# Dissimilarity to control over time

```{r}
head(cover)
x = cover %>% filter(block == 1 & sp.trt == "Natives" & subblock == 1)

dis_to_control <- function(x){
  commat = x %>% select_if(is.numeric)
  euc_dis = as.matrix(vegdist(commat, diag = TRUE, method = "euclidean"))[x$year == 2008,]
  bc_dis = as.matrix(vegdist(commat, diag = TRUE, method = "bray"))[x$year == 2008,]
  year = x$year
  return(data.frame(year, bray = bc_dis, euc = euc_dis))
}

dis.dat = cover %>% 
  filter(fertilization == "none" & clip == "none" & water != "control-2nd") %>%
  group_by(block, subblock, sp.trt, water) %>% 
  do(dis_to_control(.))

# Define plotting function
plot_fgs <- function(water){
  
  p = dis.dat %>% filter(water == water) %>%
    ggplot(aes(x = as.numeric(year),
               y = bray,
               color = sp.trt)) +
    geom_point(alpha = 0.25) + 
    stat_smooth(se = FALSE) + 
    ylim(0,1) +
    ggtitle(paste(water, "H2O")) +
    scale_x_continuous(breaks = seq(0, 10, by = 2), 
                     labels = as.character(seq(2008,2018, by = 2))) +
    ylab("Dissimilarity to State in First Year") +
    xlab("Year")
  
  return(p)
}

for(i in unique(dis.dat$water)){
  p = plot_fgs(i)  
  plot(p)
}
```

```{r}
head(cover)
x = cover %>% filter(block == 1 & sp.trt == "Natives" & subblock == 1)

dis_to_prev <- function(x){
  commat = x %>% select_if(is.numeric)
  euc_dis = as.matrix(vegdist(commat, diag = TRUE, method = "euclidean"))[matrix(c(2:(nrow(commat)),(1:(nrow(commat) - 1))), ncol = 2)]
  
  
  bc_dis = as.matrix(vegdist(commat, diag = TRUE, method = "bray"))[matrix(c(2:(nrow(commat)),(1:(nrow(commat) - 1))), ncol = 2)]
  year = x$year[-1]
  return(data.frame(year, bray = bc_dis, euc = euc_dis))
}

dis.dat = cover %>% 
  filter(fertilization == "none" & clip == "none" & water != "control-2nd") %>%
  group_by(block, subblock, sp.trt, water) %>% 
  do(dis_to_prev(.))

dis.dat %>% 
  ungroup() %>%
  summarise(n = n(),
            bray_group = mean(bray),
            se_group = sd(bray) / sqrt(n))

# Define plotting function
plot_fgs <- function(water){
  
  p = dis.dat %>% filter(water == water) %>%
    ggplot(aes(x = as.numeric(year),
               y = bray,
               color = sp.trt)) +
    geom_point(alpha = 0.25) + 
    stat_smooth(se = FALSE) + 
    ylim(0,1) +
    ggtitle(paste(water, "H2O")) +
    scale_x_continuous(breaks = seq(0, 10, by = 2), 
                     labels = as.character(seq(2008,2018, by = 2))) +
    ylab("Dissimilarity to State in First Year") +
    xlab("Year")
  
  return(p)
}

for(i in unique(dis.dat$water)){
  p = plot_fgs(i)  
  plot(p)
}
```


```{r}
head(cover)
x = cover %>% filter(plot == 1)

dis_to_control <- function(x){
  commat = x %>% select_if(is.numeric)
  colnames(dist(as.numeric(x$year)))
  euc_dis = as.matrix(vegdist(commat, diag = TRUE, method = "euclidean")[dist(as.numeric(x$year)) == 1])
  bc_dis = as.matrix(vegdist(commat, diag = TRUE, method = "bray")[dist(as.numeric(x$year)) == 1])
  year =  x$year
  
  return(data.frame(start = (year[(length(year) - length(euc_dis) + 1):length(year)]), 
                    bray = bc_dis, 
                    euc = euc_dis))
}

dis.dat = cover %>% 
  filter(fertilization == "none" & clip == "none" & h2o != "control-2nd") %>%
  group_by(plot, h2o, sp.trt) %>% 
  do(dis_to_control(.))

# Define plotting function
plot_fgs <- function(water){
  p = dis.dat %>% filter(h2o == water) %>%
    ggplot(aes(x = as.numeric(start),
               y = bray,
               color = sp.trt)) +
    geom_point(alpha = 0.25) + 
    stat_smooth(se = FALSE) + 
    ylim(0,1) +
    ggtitle(paste(water, "H2O")) +
    scale_x_continuous(breaks = seq(2008,2018, by = 2), 
                     labels = as.character(seq(2008,2018, by = 2))) +
    ylab("Distance to Block Centroid (sqrt Bray-Curtis)") +
    xlab("Year")
  
  return(p)
}

for(i in unique(dis.dat$h2o)){
  p = plot_fgs(i)  
  plot(p)
}

```
