---
title: "R Notebook"
output:
  html_document:
    df_print: paged
    theme: journal
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Cleaning up WAPS data
library(dplyr);library(tidyr);library(vegan);library(testthat)

setwd("C:/Users/ebatz/Box Sync/Eviner lab shared/Evan/Research Projects/WAPS Data")

waps.comp <- read.csv("C:/Users/ebatz/Box Sync/Eviner lab shared/Evan/Research Projects/WAPS Data/WAPSthrough2017.csv",
          header = TRUE,
          stringsAsFactors = FALSE)

trtlabels <- read.csv("C:/Users/ebatz/Box Sync/Eviner lab shared/Evan/Research Projects/WAPS Data/trtlabels.csv",
          header = TRUE,
          stringsAsFactors = FALSE)

waps.specs <- read.csv("C:/Users/ebatz/Box Sync/Eviner lab shared/Evan/Research Projects/WAPS Data/WAPSspecies.csv", 
                       stringsAsFactors = F)
```

### Data cleaning

There are a bunch of small inconsistencies that come with manual data entry that need to be corrected prior to analysis.

__Checklist:__  

* Convert dataset to wide (matrix format) __done__

* Standardize species names __done__

* Clean up year column __done__

* Assign NA cover values to 0 __done__

* Standardize treatment labels __done__

* Add "class" statistics to attribute table __done__

```{r,  echo = FALSE, message = FALSE, warning = FALSE}
# Gathers dataset to long format before cleaning up, creates species column
long.comp <- gather(waps.comp, key = "species", value = "cover", -c(1:7))

# Removes year labels from species column
spec <- gsub("X\\d\\d\\.*", "", long.comp$species)

# Grabs the first two digits from the species column to make a "year" column
yr <- gsub('\\.', "", long.comp$species)
yr <- gsub("X", "20", yr)
yr <- gsub("[[:alpha:]]+0?", "", yr)

# Reassigns species variable
long.comp$species <- spec

# Attaches year variable
long.comp <- cbind(long.comp, year = as.numeric(yr))

# Lowercase names for dataset
names(long.comp) <- c("plot", "block", "subblock",
                      "water", "spcomp", "fertilization",
                      "clipping", "species", "cover", "year")

# Assigns all NA cover values to zero
long.comp$cover[is.na(long.comp$cover)] <- 0
```

```{r,  echo = FALSE, message = FALSE, warning = FALSE}
# Reassigns species names to standardized names in the waps.specs dataset
for(i in waps.specs$Species){
  long.comp$species <- gsub(paste(i, "$", sep = ""), waps.specs$Newval[waps.specs$Species == i], long.comp$species)
}

# Removes spaces from species names
long.comp$spcomp <- gsub("[[:space:]]\\+[[:space:]]*", " \\+", long.comp$spcomp)

# Removes whitespace from treatment names
long.comp$spcomp <- gsub("[[:space:]]", "", long.comp$spcomp)
long.comp$water <- gsub("[[:space:]]", "", long.comp$water)

# Assigns all NAs to zero, merges duplicated species names
long.comp$cover <- as.numeric(long.comp$cover)
long.comp$cover[is.na(long.comp$cover)] <- 0
long.comp <- long.comp %>% group_by(plot, block, subblock, water, fertilization, species, year) %>% summarise(cover = sum(cover))

# Spreads dataset to wide format
comp.wide <- spread(long.comp, key = "species", value = "cover")
comp.attr <- select(comp.wide, one_of(c("year", waps.specs$Newval[waps.specs$Type == "class"])))
comp.wide <- ungroup(comp.wide) %>%  select(one_of(waps.specs$Newval[waps.specs$Type != "class"]))
comp.wide <- apply(comp.wide, 2, as.numeric)
comp.wide[is.na(comp.wide)] <- 0

# Merges with treatment labels
comp.attr <- left_join(comp.attr, trtlabels)
comp.attr$sp.trt <- gsub("[[:space:]]", "", comp.attr$sp.trt)
```

After cleaning, we now have two separate dataframes, one that corresponds to different plot treatments/labels and another that corresponds to species abundances:

```{r}
head(comp.attr)[,1:10]
head(comp.wide)[,1:10]
```

The species matrix contains a total of 50 unique species. The full list is provided below:
```{r}
colnames(comp.wide)
```


# Ordination analysis

```{r,  echo = FALSE, message = FALSE, warning = FALSE}
onegroup_trts = c("annuals", "WAPS", "Natives")
multigroup_trts = c("natives+WAPS+annuals", "annuals+natives", "WAPS+annuals", "WAPS+natives", "Centaureasolstitialis+WAPS+annuals")
onespec_trts = unique(comp.attr$sp.trt)[!unique(comp.attr$sp.trt) %in% c(onegroup_trts, multigroup_trts)]

selected.trts <- c(onegroup_trts)
selected.clips <- c("none")
selected.fert <- c("none")
selected.water <- c("control-2nd") 

ord.mat <- comp.wide[comp.attr$water == selected.water &
                     comp.attr$sp.trt %in% selected.trts &
                     comp.attr$clip %in% selected.clips&
                     comp.attr$fert %in% selected.fert,]

ord.att <- comp.attr[comp.attr$water == selected.water &
                     comp.attr$sp.trt %in% selected.trts &
                     comp.attr$clip %in% selected.clips &
                     comp.attr$fert %in% selected.fert,]

expect_true(nrow(ord.att) == nrow(ord.mat))

# A number of rows have no recorded live plants (not s)
ord.att <- ord.att[rowSums(ord.mat) != 0, ]
ord.mat <- ord.mat[rowSums(ord.mat) != 0, ]
ord.mat <- sqrt(ord.mat)

distmat <- vegdist(ord.mat)
WAPS.mds <- metaMDS(distmat, try = 5, trymax = 50, trace = F)
MDS_xy <- cbind(data.frame(ord.att), data.frame(X = WAPS.mds$points[,1], Y = WAPS.mds$points[,2]))

#WAPS.pca <- prcomp(ord.mat)
#PCA_xy <- cbind(data.frame(ord.att), data.frame(X = WAPS.pca$x[,1], Y = WAPS.pca$x[,2]))
```

Subsetting these points over time can show us how communities change in their compositional similarity, both between each other and within each group. The immediate pattern that I see is that communities appear to collapse togather eventually. 
```{r}
library(ggplot2)
ggplot(MDS_xy, aes(X, Y, color = sp.trt, shape = clip)) + geom_point() + theme_bw() + facet_wrap(~year)
```

```{r}
ggplot(MDS_xy[order(MDS_xy$year),], aes(X, Y, color = as.factor(plot))) + 
  geom_point() + 
  geom_path( arrow = arrow(unit(0.30,"cm"), angle = 15, ends = "last", type = "closed"), alpha = .5) + 
  theme_bw() + 
  facet_grid(sp.trt ~ block) +
  guides(color = FALSE)

```
```{r}
group_abund <- MDS_xy %>% gather(key = "group", value = "cover", annuals, bare, waps, natives, weeds) %>%
  replace_na(list(cover = 0)) 

ggplot(data = group_abund, 
       aes(x = year,
           y = cover,
           color = group)) +
  geom_point() + 
  facet_wrap(~sp.trt) +
  stat_smooth(method = "lm")
```


```{r}
calcRateChange_bray <- function(comdf, comvar, abundvar, timevar){
  # Take dataframe,
  # Cast by species variable and abundance
  # Take just numeric values
  # Assign to commat
  ssdf <- comdf[,colnames(comdf) %in% c(comvar, abundvar, timevar)]
  commat <- spread(data.frame(ssdf), 
                   comvar,
                   abundvar,
                   drop = TRUE,
                   fill = 0)
  
  # Calculate vegetation distance
  vd <- vegdist(commat[,-1], method = "euclidean")
  td <- dist(commat[,colnames(commat) == timevar])
  output <- cbind(interval = c(td), distance = c(vd))
  return(data.frame(output))
}

comp_ratechange <- left_join(long.comp, trtlabels)

comp_ratechange <- comp_ratechange  %>% filter(species %in% waps.specs$Newval[waps.specs$Type != "class"] &
                                        water == "control-2nd" &
                                        sp.trt %in% selected.trts &
                                        clip %in% selected.clips &
                                        fert %in% selected.fert)


temp_ratechange_bray <- data.frame(comp_ratechange) %>% group_by(block, plot, sp.trt, water) %>%
  filter(sp.trt %in% onegroup_trts) %>%
  filter(cover > 0) %>%
  mutate(cover = cover) %>%
  do(calcRateChange_bray(., timevar = "year",
                         comvar = "species",
                         abundvar = "cover"))

ggplot(aes(x = interval,
           y = distance,
           color = sp.trt),
       data = temp_ratechange_bray) +
  geom_point(alpha = .1) +
  facet_wrap(~sp.trt, scales = "fixed") + 
  stat_smooth(method = "lm", se = FALSE)

```
```{r}
summary(lm(distance ~ interval * sp.trt, data = temp_ratechange_bray))

plot(lm(distance ~ interval * sp.trt, data = temp_ratechange_bray))
```


